# COVID-19 Variant Analysis: Spike Protein Mutations

This analysis compares the spike protein sequences of different COVID-19 variants against the original Wuhan reference strain to identify key mutations.

## Setup and Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 8)
```

```{r libraries}
# Load required libraries for sequence analysis and visualization
library(seqinr)      # DNA/protein sequence analysis
library(dplyr)       # Data manipulation
library(ggplot2)     # Data visualization
library(Biostrings)  # Bioconductor sequence handling
library(pwalign)     # Pairwise sequence alignment
library(foreach)     # Parallel processing
library(future)      # Parallel backend
library(doFuture)    # Future adapter for foreach

# Setup parallel processing
plan(multicore)      # More stable than multicore for Biostrings
registerDoFuture()

# Set random seed for reproducibility
set.seed(123)
```

## File Configuration

```{r file_paths}
# Define file paths for reference and variant sequences
ref_file <- "wuhan.fasta"
variant_files <- c(
  "alpha.fasta",
  "beta.fasta", 
  "gamma.fasta",
  "delta.fasta",
  "omicronba1.fasta",
  "omicronba2.fasta",
  "omicronjn1.fasta"
)

# Create variant names for labeling
variant_names <- gsub(".fasta$", "", basename(variant_files))
```

## Helper Functions

### Sequence Cleaning Function
This function removes problematic characters from DNA sequences, replacing them with 'A' to maintain sequence length.

```{r clean_function}
clean_seq <- function(seq) {
  # Replace any non-standard nucleotides (not A, C, T, G, U) with 'A'
  gsub("[^ACTGUactgu]", "A", seq)
}
```

### Surface Glycoprotein Extraction
Extracts spike protein (surface glycoprotein) sequences from FASTA files and limits to first 10 sequences per variant.

```{r extract_function}
extract_surface_glycoproteins <- function(filepath, max_seqs = 10) {
  # Read FASTA file
  fasta_seqs <- read.fasta(filepath, seqtype = "DNA", as.string = TRUE, whole.header = TRUE)
  headers <- names(fasta_seqs)
  
  # Find sequences containing "surface glycoprotein" in header
  is_surface <- grepl("\\|surface glycoprotein", headers)
  
  if (sum(is_surface) == 0) {
    stop("No surface glycoprotein genes found in the file")
  }
  
  # Extract surface glycoprotein sequences
  surface_seqs <- fasta_seqs[is_surface]
  
  # Limit to first max_seqs sequences
  if (length(surface_seqs) > max_seqs) {
    surface_seqs <- surface_seqs[1:max_seqs]
    cat(sprintf("Limited to first %d sequences from %s\n", max_seqs, basename(filepath)))
  }
  
  sequences <- unlist(surface_seqs)
  names(sequences) <- names(surface_seqs)
  return(sequences)
}
```

### Sequence Alignment Function
Performs global pairwise alignment using the Needleman-Wunsch algorithm.

```{r alignment_function}
needleman_wunsch <- function(ref_seq, var_seq) {
  # Clean and convert sequences to DNAString objects
  ref_seq <- DNAString(clean_seq(ref_seq))
  var_seq <- DNAString(clean_seq(var_seq))
  
  # Define nucleotide substitution matrix
  dna_mat <- nucleotideSubstitutionMatrix(match = 1, mismatch = -1, baseOnly = FALSE)
  
  # Perform global pairwise alignment
  alignment <- pairwiseAlignment(
    pattern = var_seq,
    subject = ref_seq,
    substitutionMatrix = dna_mat,
    gapOpening = -0.5,
    gapExtension = -0.5,
    type = "global"
  )
  
  # Extract aligned sequences
  aln_ref <- as.character(subject(alignment))
  aln_var <- as.character(pattern(alignment))
  
  return(c(aln_ref, aln_var))
}
```

### Mutation Detection Function
Compares aligned sequences codon by codon to identify amino acid changes. **Fixed mutation position calculation**.

```{r comparison_function}
compare_sequences <- function(ref_seq, var_seq, variant_name) {
  # Align sequences
  aligned <- needleman_wunsch(ref_seq, var_seq)
  aln_ref <- aligned[1]
  aln_var <- aligned[2]
  
  # Ensure sequence length is divisible by 3 (complete codons)
  len <- min(nchar(aln_ref), nchar(aln_var))
  len <- len - (len %% 3)
  
  result <- list()
  idx <- 0
  
  # Compare sequences codon by codon
  for (i in seq(1, len, by = 3)) {
    ref_codon <- substr(aln_ref, i, i + 2)
    var_codon <- substr(aln_var, i, i + 2)
    
    # Skip if codons contain gaps or are identical
    if (grepl("-", ref_codon) | grepl("-", var_codon) | ref_codon == var_codon) next
    
    # Translate codons to amino acids
    ref_aa <- seqinr::translate(s2c(ref_codon))
    var_aa <- seqinr::translate(s2c(var_codon))
    
    # Determine mutation type
    mut_type <- ifelse(ref_aa == var_aa, "synonymous", "non-synonymous")
    
    # FIXED: Calculate proper amino acid position
    aa_position <- ceiling(i / 3)  # Use ceiling to get integer position
    
    idx <- idx + 1
    result[[idx]] <- data.frame(
      aa_position = aa_position,
      ref_codon = ref_codon,
      var_codon = var_codon,
      ref_aa = ref_aa,
      var_aa = var_aa,
      aa_change = paste0(ref_aa, aa_position, var_aa),  # Now uses integer position
      type = mut_type,
      variant = variant_name,
      stringsAsFactors = FALSE
    )
  }
  
  if (idx == 0) return(NULL)
  return(bind_rows(result))
}
```

## Data Processing

### Load Reference Sequence
```{r load_reference}
# Load Wuhan reference spike protein sequence
cat("Loading reference sequence...\n")
ref_seq <- extract_surface_glycoproteins(ref_file, max_seqs = 1)[[1]]
cat(sprintf("Reference sequence loaded: %d nucleotides\n", nchar(ref_seq)))
```

### Analyze All Variants
```{r analyze_variants}
# Initialize storage for mutations
mutations_list <- list()

cat("Analyzing variants...\n")
for (file in variant_files) {
  variant_name <- gsub(".fasta$", "", basename(file))
  cat(sprintf("Processing %s...\n", variant_name))
  
  # Load variant sequences (first 10 only)
  variant_data <- extract_surface_glycoproteins(file, max_seqs = 10)
  
  # Compare each variant sequence against reference using parallel processing
  variant_mutations <- foreach(i = seq_along(variant_data), .combine = rbind) %dofuture% {
    var_seq <- variant_data[[i]]
    compare_sequences(ref_seq, var_seq, variant_name)
  }
  
  # Store mutations for this variant
  if (!is.null(variant_mutations) && nrow(variant_mutations) > 0) {
    mutations_list[[variant_name]] <- variant_mutations
  }
  
  cat(sprintf("  Completed analysis for %s\n", variant_name))
}

# Combine all mutation data
mutations <- bind_rows(mutations_list)
cat(sprintf("\nTotal mutations identified: %d\n", nrow(mutations)))
```

## Results Summary

### Mutation Overview
```{r mutation_summary}
# Summary statistics
summary_stats <- mutations %>%
  group_by(variant, type) %>%
  summarise(count = n(), .groups = "drop")

# Create summary table manually
variant_summary <- summary_stats %>%
  group_by(variant) %>%
  summarise(
    nonsynonymous = sum(count[type == "non-synonymous"]),
    synonymous = sum(count[type == "synonymous"]),
    total = sum(count),
    nonsynonymous_pct = round((nonsynonymous / total) * 100, 1),
    .groups = "drop"
  )

print(variant_summary)
```

### Top Mutations by Variant
```{r top_mutations}
# Find most frequent mutations per variant
top_mutations <- mutations %>%
  filter(type == "non-synonymous") %>%  # Focus on non-synonymous mutations
  group_by(variant, aa_change) %>%
  summarise(frequency = n(), .groups = "drop") %>%
  group_by(variant) %>%
  slice_max(order_by = frequency, n = 8) %>%  # Top 8 for cleaner visualization
  ungroup() %>%
  arrange(variant, desc(frequency))

# Display top mutations table
top_mutations %>%
  group_by(variant) %>%
  slice_head(n = 5) %>%
  print()
```

## Visualization

### Enhanced Mutation Frequency Plot
```{r mutation_plot, fig.width=14, fig.height=10}
# Create an enhanced visualization
p1 <- ggplot(top_mutations, aes(x = reorder(aa_change, frequency), 
                                y = frequency, fill = variant)) +
  geom_col(alpha = 0.8, width = 0.7) +
  coord_flip() +
  facet_wrap(~variant, scales = "free", ncol = 2) +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  labs(
    title = "Most Frequent Non-synonymous Spike Protein Mutations by COVID-19 Variant",
    subtitle = "Analysis of first 10 sequences per variant compared to Wuhan reference",
    x = "Amino Acid Change",
    y = "Frequency",
    caption = "Format: [Original AA][Position][New AA] (e.g., D614G = Aspartic acid → Glycine at position 614)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    strip.text = element_text(size = 11, face = "bold"),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    plot.caption = element_text(size = 9, color = "gray50")
  )

print(p1)
```

### Mutation Position Distribution
```{r position_plot, fig.width=12, fig.height=6}
# Plot showing mutation positions across the spike protein
p2 <- mutations %>%
  filter(type == "non-synonymous") %>%
  ggplot(aes(x = aa_position, fill = variant)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  facet_wrap(~variant, ncol = 2) +
  scale_fill_viridis_d(option = "plasma") +
  labs(
    title = "Distribution of Non-synonymous Mutations Along Spike Protein",
    x = "Amino Acid Position",
    y = "Number of Mutations",
    subtitle = "Shows where mutations cluster in the spike protein sequence"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )

print(p2)
```

## Key Findings

```{r key_findings}
# Identify shared mutations across variants
shared_mutations <- mutations %>%
  filter(type == "non-synonymous") %>%
  group_by(aa_change) %>%
  summarise(
    variants = paste(unique(variant), collapse = ", "),
    variant_count = n_distinct(variant),
    total_frequency = n(),
    .groups = "drop"
  ) %>%
  filter(variant_count >= 3) %>%  # Mutations in 3+ variants
  arrange(desc(variant_count), desc(total_frequency))

cat("## Mutations Found in Multiple Variants\n")
if (nrow(shared_mutations) > 0) {
  print(head(shared_mutations, 10))
} else {
  cat("No mutations found in 3 or more variants.\n")
}
```

## Interactive Mutation Explorer

```{r simple_table}
# Create simple table for detailed exploration
mutations %>%
  filter(type == "non-synonymous") %>%
  select(variant, aa_change, aa_position, ref_aa, var_aa) %>%
  arrange(variant, aa_position) %>%
  head(50) %>%  # Show first 50 mutations
  print()
```

---

**Analysis Notes:**
- This analysis compares the **first 10 spike protein sequences** from each variant file
- Mutations are identified through global pairwise alignment (Needleman-Wunsch algorithm)
- Only **non-synonymous mutations** (amino acid changes) are highlighted in visualizations
- Mutation positions are now correctly calculated as integers (fixed from previous fractional positions)
- The D614G mutation, if present, is a well-known mutation that increases viral transmission